A family of accessibility measures

### 2024-12-30 paper/manuscript.qmd
some updates to the manuscript

<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy. This has had problematic consequences. Conceptually, accessibility is a more complete and comprehensive evolution of mobility. As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs--> 
Accessibility, a central concept in transport geography, planning, and engineering, is conceptually related to human mobility and the opportunity landscape. In simple terms, it is defined as the "potential of opportunities for interaction" [@hansen1959]. The range of opportunities considered in accessibility analysis has grown to include <!-- --> employment [REFS], health care [REFS], green spaces [REFS], schools [REFS], social contacts [REFS], emergency services [e.g., shelters, REFS], and many more. In other words, accessibility analysis is today used to understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020]. Compared to earlier measures of performance used in transportation (e.g., VKT, PKT, etc.) that benchmark the amount of movement, accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997]. However, the focus of transportation planning has historically prioritized mobility and treated accessibility as a by-product of movement, leading to problematic consequences [@handy2020]. With the rise of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013], recent decades of mobility-based practice (e.g., road and highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost), often with limited impact on accessibility levels [e.g., @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010]. In response, transportation planning practice has begin to adopt accessibility planning approaches, via accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling]. However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and interpretability and communicability of the outputs [@geurs2004; @ferreiraReenactingMobilityAccessibility2020].

@@

## The inquiry into spatial interaction traces to Newton's Universal Gravitation

The patterns of people's movement in space have been a subject of scientific inquiry for centuries. From as far back as Henry C. Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858] there was already a concern with the scientific study of human interaction over space. It is in this work where Carey states that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp. 37-38], which shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-phys-grav-prop].

@@

<!-- SUMMARY: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy. -->
Following Carey's _Principles_ of 1858, research into human spatial interaction continued in different contexts. In the late 1880s Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889]. In these works, Ravenstein posited a directly proportional relationship of migration flows with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination. These propositions are similar to Newton's gravitational laws. Over time, other researchers discovered similar relationships. <!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations. Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946. The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.

@@

\noindent where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and  $D_{12}$ is the distance between them.

@@

Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic. It had issues with inconsistent mathematical notation. For example, e.g., in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant for both _demographic force_ and  _demographic energy_ on p. 34 <!--Antonio, is this what you meant?-->. Perhaps more seriously, Stewart's work was permeated by a view of humans as particles following physical laws, but with ideas that were afflicted by unscientific racism. For instance, the molecular weight $\mu$ was presumed to be one for the average American, but "presumably...much less than one....for an Australian aborigine" [p. 35]. Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the emerging field of accessibility research.

@@

## Hansen's gravity-based accessibility
<!-- SUMMARY: in Hansen's use of @stewartDemographicGravitationEvidence1948's Population Potential measure, we highlight two crucial omissions. 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it. 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient. In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed. -->

@@

## From Hansen-type accessibility to accessibility today
<!-- SUMMARY: accessibility from Hansen to today is very diverse and has a lack of concern for the interpretability of output units; likewise, a lack of concern for balancing units such as a proportionality constant.  Our answer: let's focus on those omitted (or implicit set to 1) proportionality constants.-->

Since @hansen1959, accessibility analysis has been widely employed across planning disciplines [CITE MAJOR REVIEWS OVER LAST 30 YEARS]. However, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly determining it. In this way, $G$ continues to be implicitly set to $1$ even when the fundamental relationship in accessibility is proportionality (e.g., $ S_{i}  \propto G \sum_j O_j \cdot f(d_{ij})$ and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. Without a proportionality constant, however, the units of $S_i$ remain unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.

To overview recent accessibility literature development, the majority of studies develop on; the breath and geographic case study selection of the opportunity $O_j$ studied and potentially impacted population []; the accuracy of parameters that capture distance-decay $f(c_{ij})$ []; and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024]. Overall, due to the diverse range of questions accessibility measures have been applied to help clarify, a plethora of variations in @eq-accessibility-general are present in the literature.

However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) []. Interpretability issues plague this type of analysis, as implications beyond high-or-low access hotspots are often desired. What does the accessibility value mean? Difficultly in interpreting the output's tangible meaning contributes to resistance to its application in practice [@ferreiraReenactingMobilityAccessibility2020]. As reviewed, quantity and variations on @eq-accessibility-general also compounds the issue of interpretability across studies.

A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (or gravity), utility-, location-, and person- based measures with different components. @wuUnifyingAccess2020 demonstrate that most or all of these concepts can be unified into its main features: the relationship between travel-cost and corresponding reachable opportunities. Indeed, these authors demonstrate how previous classifications are special cases of travel-cost or opportunities constraints. From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing, the proportionality constant $G$.

This paper's focus is on the proportionality constant. We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings. To address this matter, we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971. In essence, in what follows, we reunite the accessibility indicators literature with the spatial interaction literature as developed by @wilson1971 based on entropy maximization concepts. Although Wilson's approach is based on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants: including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.

# Gravity-based accessibility's cousin: Wilson's spatial interaction model

Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used [@ortuzar_2011_modelling;@wilson1971]. However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners.

There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the spatial interaction model, which takes the following form after Wilson [-@wilson1971]:

@@

**Entropy maximising: the gravity-model as statistical averages:**
Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971]. But instead of relying on the heuristic Newtonian gravity analogy, entropy maximisation has emerged as a popular model building approach taken in the spatial interaction studies to generate the gravity model [@ortuzar_2011_modelling]. Namely, it is mathematically rigorousness and offers a sensible interpretation of the resulting model as a statistical averaging procedure [@wilson1971;@seniorGravityModellingEntropy1979].

As such, @eq-phys-gravity-model can be derived to account for additional information to include constraints that reflect known information about the system and is mathematically rigorous. The interaction variable $T_ij$ is often in the unit of trips, so additional knowledge can be the total trip flows originating at each zone $W_i^{(1)}$ and the total trip flows terminating at zone $W_j^{(2)}$. These pieces of information can constrain $T_{ij}$ as follows.

The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$. Along a similar region-wide perspective, the total flows multiplied by their travel cost in the region is constrained to equal $C$:
$$
\sum_i\sum_j T_{ij} = T
$${#eq-constraint0-gravitymodel}

@@

The entropy maximation derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967). The basis of the approach is to accept that there are macro, meso, and micro states. In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region. Using these assumptions, we aim to find a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel). From this approach, the gravity model can be interpreted to say that given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level). There is then a most probable distribution of trips between zones (meso-states) that stasis all the macro-state constraints.

Using these constraints, the gravity model @eq-phys-gravity-model is can now be distinguished into four general cases. These cases are presented in increase order of constraint restrictiveness and continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:

@@

The following are simple numerical examples of each case assuming the following 1) $T_{ij}$ trip ends matrix, 2) $C_{ij}$ travel cost matrix, and 3) an assumed travel impedance function of $f(c_{ij}) = \frac{1}{c_{ij}}$. For all cases, we look to estimate each $T_{ij}$ value using these three inputs, we present the known trip matrix and the trip ends below:
```{r}
known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
W_i <- known_trip_matrix |> rowSums()
W_j <- known_trip_matrix |> colSums()
T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA)
knowns_longer <- cbind(known_trip_matrix,W_i)
knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T))

## Unconstrained interaction case

$k$ can be understood as being set to 1, so the model looks as follows, with no constraints and the trip matrix is as follows. For example, $T_{1,1} = 160*160*\frac{1}{2}$

$$
T_{ij} = W_i^{(1)} W_j^{(2)} f(c_{ij})
$$
```{r}
T_unc <- W_i %*% t(W_j) / C

W_i_unc <- T_unc |> rowSums()
W_j_unc <- T_unc |> colSums()
T_TOTAL_unc <- ifelse(sum(W_i_unc|>round())==sum(W_j_unc|>round()), sum(T_unc), NA)
T_unc_longer <- cbind(T_unc,W_i_unc)
T_unc_longer <- rbind(T_unc_longer, cbind(t(W_j_unc),T_TOTAL_unc))

rownames(T_unc_longer) <- c("1","2","3", "TOTAL") 
colnames(T_unc_longer) <- c("1","2","3", "TOTAL")

T_unc_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

As we can surmise, the trip ends do not add up to anything we know -- they do not add up to the known trip ends. Instead, they take on the assumptions of all balancing factors $k$ being equal to 1. When would this happen? It's hard to say... as the balancing factors themselves do not have practical meaning; what has meaning are the constraints. By satisfying the constraints, the model gains meaning. In this way, the resulting unconstrained interaction matrix is simply an expression of the magnitude of interaction of $W_i$ to $W_j$ assuming $f(c_{ij})$. In the context of spatial interaction modelling, this model is not entirely useful; it is in units of 'trips' but trips that satisify the macro constraint of $k$ equaling 1, which is practically meaningless.

## Total constrained interaction case

In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows:
$$
T_{ij}' = K W_i^{(1)} W_j^{(2)} f(c_{ij})
$$
\where Solved for our example, $K$ is equal to the total number of known trips divided by the sum of all the unconstrained trips made:
$$
K = \frac{T}{\sum_{i}\sum_{j} W_i^{(1)} W_j^{(2)} f(c_{ij})}\\
K = \frac{T}{\frac{W_1^{(1)}W_1^{(2)}}{c_{1,1}}+\frac{W_2^{(1)}W_1^{(2)}}{c_{2,1}}+\frac{W_3^{(1)}W_1^{(2)}}{c_{3,1}}+\frac{W_1^{(1)}W_2^{(2)}}{c_{1,2}} +... \frac{W_3^{(1)}W_3^{(2)}}{c_{3,3}}
}\\
K = \frac{790}{166443} \\
K = 0.00474636
$$

```{r}
K <- T/sum(sum(T_unc))
T_tot <- K * T_unc

W_i_tot <- T_tot |> rowSums()
W_j_tot <- T_tot |> colSums()
T_TOTAL_tot <- ifelse(sum(W_i_tot|>round())==sum(W_j_tot|>round()), sum(T_tot), NA)
T_tot_longer <- cbind(T_tot,W_i_tot)
T_tot_longer <- rbind(T_tot_longer, cbind(t(W_j_tot),T_TOTAL_tot))

rownames(T_tot_longer) <- c("1","2","3", "TOTAL") 
colnames(T_tot_longer) <- c("1","2","3", "TOTAL")

T_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```
While we can see that the total trips `r T` results, the total trip ends are not equal to the known trip ends as these constraints are not specified and thus not satisfied by the model.

## Singly-constrained

Two versions of this case, either origin-constrained or destination-constrained.

>>>>

<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy. This has had problematic consequences. Conceptually, accessibility is a more complete and comprehensive evolution of mobility. As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs--> 
Accessibility, a central concept in transport geography, planning, and engineering, is conceptually related to human mobility and the opportunity landscape. In simple terms, it is defined as the "potential of opportunities for interaction" [@hansen1959]. The range of opportunities considered in accessibility analysis has grown to include <!-- --> employment [REFS], health care [REFS], green spaces [REFS], schools [REFS], social contacts [REFS], emergency services [e.g., shelters, REFS], and many more. In other words, accessibility analysis is today used to understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020]. Compared to earlier measures of performance used in transportation (e.g., VKT, PKT, etc.) that benchmark the amount of movement, accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997]. However, the focus of transportation planning has historically prioritized mobility and treated accessibility as a by-product of movement, leading to problematic consequences [@handy2020]. With the rise of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013], recent decades of mobility-based practice (e.g., road and highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost), often with limited impact on accessibility levels [e.g., @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010]. In response, transportation planning practice has begin to adopt accessibility planning approaches, via accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling]. However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and interpretability and communicability of the outputs [@geursAccessibilityEvaluationLanduse2004; @ferreiraReenactingMobilityAccessibility2020].

@@

## The inquiry into spatial interaction traces to Newton's Universal Gravitation
<!--SUMMARY: Overview of the proportionality constant in Newton Law of Universal Gravitation - it takes the proportion relationship into one of equality. -->
The patterns of people's movement in space have been a subject of scientific inquiry for centuries. From as far back as Henry C. Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858] there was already a concern with the scientific study of human interaction over space. It is in this work where Carey states that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp. 37-38], which shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-phys-grav-prop].

@@

<!-- SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s. These researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy. -->
Following Carey's _Principles_ of 1858, research into human spatial interaction continued in different contexts. In the late 1880s Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889]. In these works, Ravenstein posited a directly proportional relationship of migration flows with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination. These propositions are similar to Newton's gravitational laws. Over time, other researchers discovered similar relationships. <!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations. Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946. The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.

@@

\noindent Where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and  $D_{12}$ is the distance between them.

@@

Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic. It had issues with inconsistent mathematical notation. For example, e.g., in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant for both _demographic force_ and  _demographic energy_ on p. 34. Perhaps more seriously, Stewart's work was permeated by a view of humans as particles following physical laws, but with ideas that were afflicted by unscientific racism. For instance, the molecular weight $\mu$ was presumed to be one for the average American, but "presumably...much less than one....for an Australian aborigine" [p. 35]. Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the emerging field of accessibility research.

@@

## Hansen's gravity-based accessibility to today
<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up. Is now cited as the father of accessibility. However, Hansen work was applied, and he's use of Stewart's Population Potential measure included two crucial omissions that afflict the literature to this day. The first omission is 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it. The second omission is 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient. In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed. -->

@@

Since @hansen1959, accessibility analysis has been widely employed across planning disciplines [@koenigIndicatorsUrbanAccessibility1980; @geursAccessibilityEvaluationLanduse2004; @millerAccessibilityMeasurementApplication2018; @wuUnifyingAccess2020]. However, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly determining it. In this way, $G$ continues to be implicitly set to $1$ even when the fundamental relationship in accessibility is proportionality (e.g., $ S_{i}  \propto G \sum_j O_j \cdot f(d_{ij})$ and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. Without a proportionality constant, however, the units of $S_i$ remain unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.

However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) [@millerAccessibilityMeasurementApplication2018]. Interpretability issues plague this type of analysis, as implications beyond high-or-low access hotspots are often desired. What does the accessibility value mean? Difficultly in interpreting the output's tangible meaning contributes to resistance to its application in practice [@ferreiraReenactingMobilityAccessibility2020]. Furthermore, The majority of accessibility studies within the past few decades develop on the breath and geographic case study selection of the opportunity $O_j$ studied and potentially impacted population []; the accuracy of parameters that capture distance-decay $f(c_{ij})$ []; and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024]. Overall, due to the diverse range of questions accessibility measures have been applied to help clarify, a plethora of variations in @eq-accessibility-general are present in the literature. These variations also compounds the issue of interpretability across studies.

A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (or gravity), utility-, location-, and person- based measures with different components. @curtis2010planning further classifies accessibility measures into spatial separation, contour-, gravity-, competition-, time-space, utility-, and network- based measures [@curtis2010planning], though methodologies may clearly overlap. Broader approaches have also been taken: @paez2012measuring categories indicators on how they have been normatively and positively applied, and @levineCenturyEvolutionAccessibility2020 reviews accessibility's evolution along definition-based and application-based dimensions throughout the years.

Among these reviews, we find @wuUnifyingAccess2020's contribution compelling for unifying accessibility measures: they demonstrate that most or all of these concepts can be unified into its main features: the relationship between travel-cost and corresponding reachable opportunities. Indeed, these authors demonstrate how previous classifications are special cases of travel-cost or opportunities constraints. From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing, the proportionality constant $G$.

As such, this paper's focus is on the proportionality constant. We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings. To address this matter, we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971. In essence, in what follows, we reunite the accessibility indicators literature with the spatial interaction literature as developed by @wilson1971 based on entropy maximization concepts. Although Wilson's approach is based on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants: including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.

# Introducing a family of accessibility measures

## Gravity-based accessibility's cousin: Wilson's spatial interaction model

Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used [@ortuzar_2011_modelling;@wilson1971]. However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners. From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the spatial interaction model, which takes the following form after Wilson [-@wilson1971]:

@@

Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971]. But instead of relying on the heuristic Newtonian gravity analogy, entropy maximisation has emerged as a popular model building approach taken in the spatial interaction studies to generate the gravity model [@ortuzar_2011_modelling]. Namely, it is mathematically rigorousness and offers a sensible interpretation of the resulting model as a statistical averaging procedure [@wilson1971;@seniorGravityModellingEntropy1979]. As such, @eq-phys-gravity-model can be derived to account for additional information to include constraints that reflect known information about the system and is mathematically rigorous. The interaction variable $T_ij$ is often in the unit of trips, so additional knowledge can be the total trip flows originating at each zone $W_i^{(1)}$ and the total trip flows terminating at zone $W_j^{(2)}$. These pieces of information can constrain $T_{ij}$ as follows.

The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$: <!-- Along a similar region-wide perspective, the total flows multiplied by their travel cost in the region is constrained to equal $C$:-->
$$
\sum_i\sum_j T_{ij} = T
$${#eq-constraint0-gravitymodel}

@@

The entropy maximization derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967). The basis of the approach is to accept that there are macro, meso, and micro states. In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region. Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel). From this approach, the gravity model can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level). There is then a most probable distribution of trips between zones (meso-states) that stasis all the macro-state constraints.

Using these constraints, the gravity model @eq-phys-gravity-model is distinguished into four general cases. These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:

@@

## Accessibilities' development seperate from the family of gravity models

Upon detailing the gravity model, we can now discuss how accessibility has been developed within that field and how it has not. Gravity modellers (interaction variable being trips, a model of spatial interaction - mobility) continued to use accessibility within transportation demand modelling. As succinctly summarized in @morrisAccessibilityIndicatorsTransport1979, "Mobility and accessibility together influence an individual’s capacity to travel in daily life". However, the gravity model literature that focuses on Hansen's accessibility is relatively scarce. To capture this sentiment, see @fig-Fig1 which demonstrates the bibliometric coupling of papers that cites @hansen1959, @wilson1971, and both @hansen1959 and @wilson1971. @fig-Fig1 is based on a SCOPUS search of papers that cite @hansen1959's DOI (2,134 documents) and another SCOPUS search of papers that cite @wilson1971's DOI (368 documents) conducted on November 9, 2024 and a bibliometric coupling matrix created using the {bibliometrix} R package [@bibliometrix2017].

```{r plot-figure1, out.width=600, out.height=600}
#| label: fig-Fig1
#| fig-cap: "The bibliometric coupling of papers that cite Hansen (1959), Wilson (1971) and both Hansen (1959) and Wilson (1971)."
include_graphics(glue(here(), "/data-raw/Han_Wilson_bib_coupling_plot.png"))
```

Many of the papers that cite both @hansen1959 and @wilson1971 (77 journal articles) are transportation modelling papers working on integrating accessibility within the gravity model (mobility).

For instance, @battyMethodResiduesUrban1976 states that the relationship between the gravity model and Hansen's original accessibility-potential model may be worth exploring... There is a consistent acknowledgement that these two models are related. Specifically, that the Hansen-type accessibility model is embedded within the balancing factors [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985]. This is a "common sense" approach at integrating accessibility [@morrisAccessibilityIndicatorsTransport1979].

Though there are other approaches to consider accessibility with spatial interaction modelling beyond Hansen founded on micro-economic consumer behaviour [@morrisAccessibilityIndicatorsTransport1979]. For instance, others explore entropy maximising solutions (Wilson's framework and hence the balancing factors defined as Hansen's accessibility) when considering demand behavior described by random utility choice models [@leonardiRandomUtilityDemand1984].

Some work on location-allocation models also used both Hansen and Wilson. For instance, @leonardiOptimumFacilityLocation1978 and @beaumontLocationallocationProblemsPlane1981 define a location-allocation problem using the singly-constrained gravity model, and then defines one using Hansen-type accessibility as the objective function subject.

Then, @fotheringhamSpatialCompetitionAgglomeration1985 demonstrated that destination's accessibility should be explicitly specified within Wilson's framework, citing Hansen measure as convenient formulation. Without the consideration for accessibility explicitly, unexplained spatial pattern remains. This later came to be the 'competition destination' model, and is one of a few techniques to address spatial variation in gravity models. [@oshanSpatialStructureDebate2021]. Implicitly, we gather that at this point: it was understood balancing factors in the gravity model could not be understood as capturing accessibility. We also gather that in this moment of growing complexity of spatial interaction modelling, accessibility branched off and sprouted in the medium of budding GIS technology and the emergence of space-time dimensions [@millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting]. With the possibility for showcasing complex GIS methodologies for varied case studies, accessibility became mature enough to branch off on its own.

In papers after @fotheringhamSpatialCompetitionAgglomeration1985, Hansen is typically cited as providing a definition of accessibility but how Wilson is cited changes through the years. Wilson is attributed for using different travel cost functions in depending on the context [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].

Othertimes, Wilson and Hansen measures were progressively incorrectly conflated throughout the years [ @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring] - indeed they both measure spatial interaction so are somewhat related, but they are not the same. They are also co-cited to refer to the concept of 'gravity' in models [@liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending]

And finally, Wilson's balancing factors are interpreted as inverse Hansen-type accessibility, especially in the context of competition [@vickermanAccessibilityAttractionPotential1974a; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. This interpretation is fraut, upon considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration, which many of these works do not contend with.

THE SPATIAL STRUCTURE DEBATE: @oshanSpatialStructureDebate2021
